---
title: "Linear Modeling Lab 1"
author: "Sports Camp Crew"
date: "May 27, 2019"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: paper
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Linear modeling

## Goals

We will briefly review linear modeling, focusing on building and assessing linear models in R. We have five main goals in this lab:

+ Use exploratory data analysis (EDA) and visualization to determine a) whether two variables have a linear relationship, and b) among a set of explanatory variables, which one(s) seem like the best candidates for
predicting a given output variable.

+ Fit a few simple multivariate linear models.

+ Interpret our fitted linear models.

+ Assess our fitted linear models.

+ Use our fitted linear models to perform prediction.

## Preliminaries

Execute the following code chunk to load the necessary packages for this lab:

```{r, echo = FALSE}
library('tidyverse')
library('GGally')
```

Execute the following code chunk to a) load the necessary data for this lab, b) compute a few variables we will use in this lab, and c) subset out players with low minute totals (fewer than 250 minutes played in a season):

```{r}
nba <- read.csv("https://raw.githubusercontent.com/ryurko/CMSACamp/master/data/intro_r/nba_2019_player_stats.csv?token=AFKSV7EZOGYYTLAWNZ7YRPC453VNQ")

nba$field_goal_percentage <- nba$field_goals / nba$field_goal_attempts
nba$three_point_percentage <- nba$three_pointers / nba$three_point_attempts
nba$free_throw_percentage <- nba$free_throws / nba$free_throw_attempts
nba_subset <- filter_all(nba, all_vars(!is.na(.))) %>% 
  filter(minutes_played > 250) 
```

## Which players play the most minutes?

In the National Basketball Association (NBA), and more generally in team sports, a coach must make decisions about how many minutes each player should play. Typically, these decisions are informed by a player's skills, along with other factors such as fatigue, matchups, etc. Our goal today is to use measurements of a few (quantifiable) player attributes to predict the number of minutes a player plays. In particular, we have the following data, measured over the 2019 NBA regular season for 439 total players:

+ `player`: The names of each player (not useful for modeling purposes, but just for fun!)
+ `minutes_played`: Our response variable, measuring the total number of minutes a player played over the entire 2019 NBA regular season.
+ `field_goal_percentage`: Potential (continuous) explanatory variable, calculated as (number of made field goals) / (number of field goals attempted).
+ `free_throw_percentage`: Potential (continuous) explanatory variable, calculated as (number of made free throws) / (number of free throws attempted).
+ `three_point_percentage`: Potential (continuous) explanatory variable, calculated as (number of made 3 point shots) / (number of 3 point shots attempted).
+ `position`: Potential (categorical) explanatory variable, one of `SG` (shooting guard), `PG` (point guard), `C` (center), `PF` (power forward) or `SF` (small forward).

## Exploratory Data Analysis

To start with, let's do some EDA to visually assess which of our explanatory variables is most associated with number of minutes played. Execute the following code chunk to make a scatterplot showing the relationship of field goal percentage with number of minutes played:

```{r}
ggplot(data = nba_subset, aes(x = field_goal_percentage, y = minutes_played)) + geom_point()
```

**Question 1**. Describe the direction and strength of the association between field goal percentage and minutes played. Does the relationship between the explanatory variable and response variable appear to be linear?

**Question 2**. Now, in the following code chunks, write code making equivalent scatterplots showing the relationships between free throw percentage and minutes played, and three point percentage and minutes played.

```{r}
```

```{r}
```

Using the two scatterplots you've just built, describe the direction and strength of the association between the explanatory and response variables. In each case, does the relationship between the explanatory variable and response variable appear to be linear?

**Question 3**. In your opinion, which of the three continuous explanatory variables (field goal percentage, free throw percentage, three point percentage) is the best predictor of minutes played?

**Question 4**. Build a box plot showing the relationship between position and minutes played. Does there appear to be a relationship?


## Fit a linear model

Now that we've performed some EDA, it's time to actually fit some linear models to our data. Execute the following code chunk to a) fit a linear model predicting minutes played as a function of field goal percentage and position, and b) summarize the fitted model. 

```{r}
fieldgoalpos_linmod <- lm(minutes_played ~ field_goal_percentage + position, data = nba_subset)
summary(fieldgoalpos_linmod)
```

**Question 1**. In the following code chunks, fit and store two other linear models, predicting minutes played using two different subsets of the explanatory variables.

```{r}
```

```{r}
```

**Question 2**. In words, interpret the intercept and slope coefficients for each of the three models we've just fitted. 

## Assess fitted models

Now that we have three candidate linear models for predicting minutes played, we need to determine which model we prefer. We can do so either visually or numerically.

**Question 1**. The `summary` function, when called on a stored linear model, computes a bunch of interesting information about that model. One such interesting statistic is known as the $R^2$, which measures the proportion of variation in the response variable explained by the explanatory variable. In simple terms, the higher the $R^2$ of a one-variable linear model, the more useful the explanatory variable was in predicting the response variable (using a linear model).

Execute the following code chunk to compute the $R^2$ of the linear model predicting minutes played using field goal percentage.

```{r}
fieldgoalpos_rsq <- summary(fieldgoalpos_linmod)$r.squared
fieldgoalpos_rsq
```

In the following code chunks, compute the $R^2$ of the other two linear models you fitted.

```{r}
```

```{r}
```

Based on $R^2$, which explanatory variable was most useful in predicting minutes played (using a linear model)? What is potentially dangerous about using $R^2$ to assess a fitted model (if you don't know, ask!)?

**Question 2**. To assess the fit of a linear model, we can also plot the predicted values vs. the actual values, to see how closely our predictions align with realit, and to decide whether our model is making any systematic errors. Execute the following code chunk to show the actual minutes played vs. the minutes played predicted using our linear model of field goal percentage + position.

```{r}
fieldgoal_preds <- data.frame(actual_y = nba_subset$minutes_played,
                              predicted_y = predict.lm(fieldgoalpos_linmod))
ggplot(data = fieldgoal_preds, aes(x = predicted_y, y = actual_y)) + 
  geom_point()
```

Use this plot to assess the linear relationship between field goal percentage and position, and minutes played.

**Question 3**. In the following code chunks, plot the scatterplots of the actual minutes played vs those predicted by the two linear models you built in the previous section.

```{r}
```

```{r}
```

**Question 4**. Which of the three linear models most accurately predicted the response variable minutes played? Use your answers to Questions 1-3 to justify your answer.

## Prediction using a linear model. 

Run the following code chunk to select the player LaMarcus Aldridge, a center for the San Antonio Spurs, and save his stats in a new data frame.

```{r}
nba_lma <- filter(nba_subset, player == 'LaMarcus Aldridge')
```

**Question 1**. Run the following code chunk to predict LaMarcus Aldridge's minutes played from his field goal percentage, using the `fieldgoal_linmod` fitted linear model.

```{r}
predict.lm(fieldgoalpos_linmod, nba_lma)
```

**Question 2**. In the following code chunks, predict LaMarcus Aldridge's minutes played from the other two linear models you fit in the previous section.

```{r}
```

```{r}
```

**Question 3**. Which model gave the most accurate prediction of LaMarcus Aldridge's minutes played?

**Note**. Clearly, it doesn't make sense to choose a model based on which did the best job of predicting LaMarcus Aldridge's minutes, as he is only one player and is unlikely to be representative of the entire NBA. As we will see later,however, the idea of choosing a model based on which model has the highest predictive accuracy is a powerful and widely applied technique. 

## Takeaways

Linear models have dramatic strengths and weaknesses. They are simple to fit and easy to interpret and explain, but they are often not a good fit for data and can be very influenced by outliers. In general, you should probably always try a linear model for a given problem, but it is unlikely that it will be the best fit for the given data. As we will see, many of the more complicated models we study can be viewed as extensions of linear models.

# Bonus Questions

**Question 1**

Fit any univariate linear model, and compare the square root of the $R^2$ of that linear model to the correlation between that explanatory variable and the response. What do you notice?

**Question 2**

Find the variable in the `nba_subset` data frame which is most correlated with minutes played. In light of the answer to Question 1, we know this must also have the highest $R^2$, and is therefore the best one-variable linear model for predicting minutes played. Despite this, why might choosing this variable not lead to the most interesting final conclusion?

**Question 3**

Execute the following code chunk to create a different subset of the nba data, where we've included some additional low minute players.

```{r}
nba_subset_2 <- filter_all(nba, all_vars(!is.na(.))) %>% 
  filter(minutes_played > 200) 
```

Refit any of your previous linear models using this new subset. How does the fitted model/ performance change?

**Question 4**

Choose an interesting pair of explanatory and response variables, and repeat the first three sections of this lab with this new pair of variables.


